---
title: "Feature Space Diagram in R"
author: "John Erickson"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This is a first effort to implement "feature space diagrams" in R, inspired by https://towardsdatascience.com/escape-the-correlation-matrix-into-feature-space-4d71c51f25e5

```{r,libraries, results=FALSE, message=FALSE, warning=FALSE}
library(lessR)
library(Hmisc)
library(corrplot)
library(gplots)
library(igraph)
library(tidyverse)
```

## Workflow

Our workflow is generally as described in the original post:

* Generate the correlation matrix
* Take the absolute value of correlation matrix and subtract each value from 1. The result is a distance matrix.
* Use PCA to reduce our NxN matrix to Nx2.
* Plot each feature’s location using the two principal components.
* Use Feature Agglomeration to generate feature clusters.
* Color each feature by its cluster.
* Draw lines to represent relationships of at least r = 0.7 (or user's choosing)

## Data Load
```{r data_load}
# TODO: User uploads data
# Load data 
boston <- read.csv("boston.csv", header = TRUE, fileEncoding="latin1")

# Ensure we have a matrix
mydata <- as.matrix(boston)
```

## Initial Correlation Matrix 

```{r correlation}
# Simple correlation matrix calculation
# TODO: user chooses method
mydata.cor <- cor(boston, method = c("pearson", "kendall", "spearman"), use = "complete.obs")

# # Optional
# corrplot(mydata.cor)
# 
# palette <- colorRampPalette(c("green", "white", "red")) (20)
# heatmap.2(x = mydata.cor, col = palette, symm = TRUE)

# Reorder correlation matrix based on: https://rdrr.io/cran/lessR/man/corReorder.html 
mydata.cor.ro <- corReorder(mydata.cor)
```

## Distance Matrix Calculation

```{r distance}
# Distance matrix (Absolute value et.al.)
mydata.cor.ro.1 <- abs(mydata.cor.ro) - 1
```

## PCA on Distance Matrix

```{r PCA} 
# PCA 
mydata.cor.ro.1.pca <- prcomp(mydata.cor.ro.1)

# Plottable version:
mydata.cor.ro.1.pca.plot <- as.data.frame(mydata.cor.ro.1.pca$x[,1:2])
```

## Finding the Clusters

```{r clusters} 
# Pull out our clusters 
mydata.cor.ro.1.pca.plot$cluster <- as.factor(kmeans(mydata.cor.ro.1.pca$x[,1:2], centers=4)$cluster)
mydata.cor.ro.1.pca.plot$name <- rownames(mydata.cor.ro.1)
```

## Finding the Graph!

This is the hard part...

```{r network}
# Determine connectivity!
# filtering
selector <- ((abs(mydata.cor.ro.1) <= 0.5 ) * 1)

for (i in 1:length(mydata.cor.ro.1.pca.plot$cluster)) {selector[,i] <- selector[,i] * as.numeric(mydata.cor.ro.1.pca.plot$cluster)}
for (i in 1:length(mydata.cor.ro.1.pca.plot$cluster)) {selector[i,i] <- 0 }

# Switching to networks
# Create igraph structure
network <- graph_from_adjacency_matrix(selector, weighted = TRUE)

# Gets us our edges!
mydata.edges <- get.data.frame(network)

# replace `from` with X1 and Y1, and `to` with X2 and Y2
# These will be our line segments!
mydata.segments <- mydata.edges %>% 
  left_join(mydata.cor.ro.1.pca.plot, by=c("from"="name")) %>%
  select(-cluster) %>%
  mutate(X1=PC1, Y1=PC2) %>%
  select(-PC1, -PC2)

mydata.segments <- mydata.segments %>%
  left_join(mydata.cor.ro.1.pca.plot, by=c("to"="name")) %>%
  select(-cluster) %>%
  mutate(X2=PC1, Y2=PC2) %>%
  select(-PC1, -PC2)

mydata.cor.ro.1.pca.plot$cluster <- as.factor(mydata.cor.ro.1.pca.plot$cluster)

# This really isn't "weight"...
mydata.segments$weight <- as.factor(mydata.segments$weight)
```

## Building the Plot

```{r plotting}
# Adding line segments to PCA plot
ggplot(mydata.cor.ro.1.pca.plot, aes(x=PC1, y=PC2)) + 
  geom_point(aes(color=cluster, size=4), show.legend = FALSE) +
  geom_text(aes(label=name),hjust=1, vjust=0) +
  geom_segment(aes(x=X1,y=Y1,xend=X2,yend=Y2,color=weight), data=mydata.segments) +
  labs(title="Feature Space Diagram for the Boston Housing data set")

# ggplot() + 
#   geom_segment(aes(x=X1,y=Y1,xend=X2,yend=Y2,color=weight), data=mydata.segments.test)

```
